Etox-Base - a tool for ecotoxicologists

### 2018-12-17 article/article.tex
changes to tikz picture and article

d\documentclass[english]{article}
\usepackage{natbib}
\usepackage{csvsimple}
\usepackage{tikz}
\usetikzlibrary{trees, arrows, shapes.geometric, positioning}
\usepackage{standalone}
%% text packages
\usepackage{siunitx} % fake ° symbol \ang{}
\usepackage[super]{nth} % for 1st, 2nd etc. \nth{}

@@

A large number of chemical compounds such as pharmaceuticals, pesticides and synthetic hormones are in daily use all over the world with insufficient knowledge about possible adverse effects on the environment. In Europe alone some 100,000 compounds are estimated to be in current use, whereof 30,000 are produced in quantities larger than one ton per year \citep{breithaupt_costs_2006}. Chemical compounds are brought to the environment deliberately - as it is the case for pesticides or arrive in the environment as a byproduct from other processes (e.g. as atmospheric emissions or as wastewater) \citep{schwarzenbach_challenge_2006}. Ecosystems provide essential services to human societies such as drinking and irrigation water, food and climate regulation. These ecosystem services are products of different ecosystem functions that crucially depend on the integrity of the populations and communities that drive these ecosystem functions.

Pollution with man-made chemical toxicants was identified as one of three major environmental problems for which research gaps hamper the derivation of planetary boundaries, i.e. thresholds beyond which irreversible state shifts may occur { | Steffen, et al., 2015 | | |zu:0:4565WXWV}.

# Überleitung Biodiversitätsverlust – Rolle von Chemikalien
Hier ein Text von mir von einem internen Antrag, vielleicht kannst Du davon was recyclen:
Ecosystems provide essential services to human societies such as drinking and irrigation water, food and climate regulation. These ecosystem services are products of different ecosystem functions that crucially depend on the integrity of the populations and communities that drive these ecosystem functions. Humans are altering ecosystems from regional (e.g. through land use change) to global (e.g. through ocean acidification) scales, which has been captured in defining the current era the Anthropocene { | Crutzen, 2002 | | |zu:0:QMD3ZN8K}. The anthropogenic alteration of ecosystems is associated with an accelerating rate of species extinctions { | Butchart, et al., 2010 | | |zu:0:PAKGEKAH}{ | Barnosky, et al., 2011 | | |zu:0:5WZP7PB6} that threaten the provisioning of the abovementioned services { | MEA, 2005 | | |zu:0:7MXCXGXS}{ | Barnosky, et al., 2014 | | |zu:0:DBW4NPGX}. The major anthropogenic stressors that have caused the deterioration of ecosystems include habitat degradation, climate change, nutrient enrichment, and chemical pollution (e.g. pesticides). Several of these stressors, such as climate change and nutrient enrichment, are close to or are already exceeding their planetary boundaries { | Steffen, et al., 2015 | | |zu:0:4565WXWV}, which may result in irreversible shifts in the biosphere { | Barnosky, et al., 2012 | | |zu:0:SXKWUBMG}. Pollution with man-made chemical toxicants was identified as one of three major environmental problems for which research gaps hamper the derivation of planetary boundaries, i.e. thresholds beyond which irreversible state shifts may occur { | Steffen, et al., 2015 | | |zu:0:4565WXWV}. Bernhardt et al. { | Bernhardt, et al., 2017 | | |zu:0:MCKMS8AZ} argue that the knowledge gap how chemicals affect biodiversity and ecosystem functions, and in turn ecosystem services, would also impede society’s ability to accomplish the Sustainable Development Goals of the United Nations. Hence, research on the effects of chemicals on ecosystems is required for establishing thresholds ensuring ecosystem integrity and enabling societies to effectively manage chemical risks. From the perspective of basic science, the scope of such a research program also involves basic scientific questions such as to which extent functions (e.g. nutrients and carbon cycling) and services are interdependent in their response to chemical stress, and to what extent non-linear feedbacks, as a potential consequence of transgenerational effects (such as induced mutation, population bottlenecks and selection ), occur.

Biodiversitätsverlust - Rolle von chemikalien?

Thereof less than one percent are thoroughly tested \citep{breithaupt_costs_2006}. For substances that have been tested adverse effects could be detected. Pesticides for exmaple reduce macro-invertebrate biodiversity in streams \citep{beketov_pesticides_2013} and estrogenic-associated effects were found in fish \citep{vethaak_integrated_2005}. However for many substance groups, especially newly emerging ones authorative test results remain scarce or not hard to find. To overcome such knowledge paucities it is paramount to accumulate and distribute existing knowledge on possible adverse effects of substances widely. The ECOTOXicology knowledgebase (ECOTOX) created by U.S. Environmental Protection Agency (EPA) represents one such initiative. It collects ecotoxicological test data on on aquatic and terrestrial wildlife as well as plants and publishes it on a quarterly basis freely accessible on the web. By the time of writing it contained 919,123 test results, comprising 11,655 chemical compounds tested on 12,630 different species \citep{elonen_ecotoxicology_2018}. The ecotoxicological tests are collected without limitation and comprise all sorts of measured endpoints such as Effective Concentrations - EC50 values, No-observed effect concentratons NOEC or lowest observed effect concentraitons - LOEC. Likewise the data set contains many different effect measures such as mortality, intoxication, growth etc. and test durations ranging from seconds to weeks or even years. These test results can be used to derive toxicity risk thresholds. As ecotoxicological risk assessment often relies on national and international quality thresholds such as EU-Environmtal quality standards (EQS) or Regulatory Aceptable concentrations (RAC) in Germany. In order to make this data more widely accessible we created a tool that downloads every new version of the \epa{} and performs several quality checks, e.g. unit harmonization, and cleaning steps on it. In addition to the test parameters that are retrieved from the \epa{} we query compound and organism parameters from other open data bases to adress missing information such as compound solubility or organism habitat or regional ocurrence patterns. Subsequently the data is provided online via a web app where users can filter the data according to needed test parameters. Furthermore the user can choose between several aggregation methods to retrieve individual values per compound. Compounds are identified by their CAS-number as this is also the key identifier in the \epa{}. Optionally the data base can also be downloaded as a whole for local useage. This is the first approach that tries to handle large amounts of toxicity test data to derive meaningful information about the susceptability of organisms towards individual compounds by aggregating test results. Researchers relying on ecotoxicological test data will benefit from the \etoxbase{} due to concise representation ecotoxicological test data.

@@

\subsection*{The application}
The application is accessible through \etox-base{} and was built by using the shiny web application framework \citep{chang_shiny_2018} in R. Therein the user can choose to filter the data by several parameters such as chemical class, organism habitat, test duration and test endpoints amongst others. A detailed list with example entries of the possible filter parameters can be found in \ref{table:meta}. In order to derive single endpoints for an organism groups and a compound of the partly extensive test data, the \etox-base{} aggregates the test results according to the chosen filters in two steps. Firstly the filtered test results are aggregated by the CAS number, the chosen taxon plus the duration and secondly only by the CAS number. The former can't be influenced by the user and calculates either the minimum or the median depending on the amount of results to aggregate (n <= 2: minimum, if n > 2: median). Thereof the second step calculates the minimum, the maximum, the median, the mean or the geometric mean as aggregate. Besides the aggregation some cleaning can also be performed. The \etox-base{} allows to exclude test results exhibiting concentrations that are higher than the actual water solubility of the respective compound at \ang{20} C. In any case the user can choose to exclude outliers. The outliers are selected by excluding values that exceed lower (0.25) and the upper (0.75) quartile by 1.5 times the interquartile range. All the steps described in this section are executed at each click in the app which runs the function fun\_ec50filter\_aggregation.R (cf. table \ref{table:processing-scripts}). Fast, adequate and live data manipulation is possible thanks to the data.table package \citep{dowle_data.table_2018}.

@@

\begin{figure}
    %\includestandalone[width=\textwidth]{article/tikz/organigram}
    \input{article/tikz/organigram}
    \caption{Organigram}
    \label{fig:organigram}
\end{figure}

@@

@@

\begin{table}
    \csvautotabular[respect all]{data/scripts.csv}
    \caption{Processing scripts}
    \label{table:processing-scripts}
\end{table}

@@

\begin{thebibliography}{}
\bibliography{references-etox-base}
\end{thebibliography}

>>>>

\documentclass[english]{article}
\usepackage{natbib} % citations
\usepackage{csvsimple}
\usepackage{rotating} % to rotate tables
%% tikz packages
\usepackage{tikz}
\usetikzlibrary{trees, arrows, shapes.geometric, positioning}
\usepackage{standalone}
%% text packages
\usepackage{siunitx} % fake ° symbol \ang{}
\usepackage[super]{nth} % for 1st, 2nd etc. \nth{}

@@

A large number of chemical compounds such as pharmaceuticals, pesticides and synthetic hormones are in daily use all over the world with insufficient knowledge about possible adverse effects on the environment. In Europe alone some 100,000 compounds are estimated to be in current use, whereof 30,000 are produced in quantities larger than one ton per year \citep{breithaupt_costs_2006}. Chemical compounds are brought to the environment deliberately - as it is the case for pesticides or arrive in the environment as a byproduct from other processes (e.g. as atmospheric emissions or as wastewater) \citep{schwarzenbach_challenge_2006}. Ecosystems provide essential services to human societies such as drinking and irrigation water, food and climate regulation. These ecosystem services are products of different ecosystem functions that crucially depend on the integrity of the populations and communities that drive these ecosystem functions. Pollution with man-made chemical toxicants threatens ecosystems in various ways which are currently not fully understood \citep{steffen_anthropocene_2007}. Besides habitat degradation, climate change and nutrient enrichment, chemical pollution poses a significant threat to earth's biodiversity

Pollution with man-made chemical toxicants was identified as one of three major environmental problems for which research gaps hamper the derivation of planetary boundaries, i.e. thresholds beyond which irreversible state shifts may occur \citep{steffen_anthropocene_2007}. Bernhardt et al. \citet{bernhardt_synthetic_2017} argue that the knowledge gap how chemicals affect biodiversity and ecosystem functions, and in turn ecosystem services, would also impede society’s ability to accomplish the Sustainable Development Goals of the United Nations. Hence, research on the effects of chemicals on ecosystems is required for establishing thresholds ensuring ecosystem integrity and enabling societies to effectively manage chemical risks.

plan:
- occurring in vast numbers
- pathways
- ecosystem services 
- biodiversity loss
- Übergang abschätzung von chemikalien

However, less than one percent are thoroughly tested \citep{breithaupt_costs_2006}. For substances that have been tested adverse effects could be detected. Pesticides for exmaple reduce macro-invertebrate biodiversity in streams \citep{beketov_pesticides_2013} and estrogenic-associated effects were found in fish \citep{vethaak_integrated_2005}. However for many substance groups, especially newly emerging ones authorative test results remain scarce or not hard to find. To overcome such knowledge paucities it is paramount to accumulate and distribute existing knowledge on possible adverse effects of substances widely. The ECOTOXicology knowledgebase (ECOTOX) created by U.S. Environmental Protection Agency (EPA) represents one such initiative. It collects ecotoxicological test data on on aquatic and terrestrial wildlife as well as plants and publishes it on a quarterly basis freely accessible on the web. By the time of writing it contained 919,123 test results, comprising 11,655 chemical compounds tested on 12,630 different species \citep{elonen_ecotoxicology_2018}. The ecotoxicological tests are collected without limitation and comprise all sorts of measured endpoints such as Effective Concentrations - EC50 values, No-observed effect concentratons NOEC or lowest observed effect concentraitons - LOEC. Likewise the data set contains many different effect measures such as mortality, intoxication, growth etc. and test durations ranging from seconds to weeks or even years. These test results can be used to derive toxicity risk thresholds. As ecotoxicological risk assessment often relies on national and international quality thresholds such as EU-Environmtal quality standards (EQS) or Regulatory Aceptable concentrations (RAC) in Germany. In order to make this data more widely accessible we created a tool that downloads every new version of the \epa{} and performs several quality checks, e.g. unit harmonization, and cleaning steps on it. In addition to the test parameters that are retrieved from the \epa{} we query compound and organism parameters from other open data bases to adress missing information such as compound solubility or organism habitat or regional ocurrence patterns. Subsequently the data is provided online via a web app where users can filter the data according to needed test parameters. Furthermore the user can choose between several aggregation methods to retrieve individual values per compound. Compounds are identified by their CAS-number as this is also the key identifier in the \epa{}. Optionally the data base can also be downloaded as a whole for local useage. This is the first approach that tries to handle large amounts of toxicity test data to derive meaningful information about the susceptability of organisms towards individual compounds by aggregating test results. Researchers relying on ecotoxicological test data will benefit from the \etoxbase{} due to concise representation ecotoxicological test data.

\citep{kefford_definition_2011} % paper that shows SSD species richness approximates effect of salinity on macroinvertebrates

\citep{posthuma_species_2002} % describes concept of SSDs first

two prominent concepts to assess effects on organisms or biodiversity in ecotoxicology are Species Sensitivity Distributions \citep{posthuma_species_2002} (SSDs) and Toxic Units (TUs) <CITATION>. Kefford et al. \citet{kefford_definition_2011} showed how SSDs to effectively approximate salinity effects on species richness (i.e. macroinvertebrates) by comparing SSDs with measurements of species richness determined in field studies.

@@

\subsection*{The application}
The application is accessible through \etoxbase{} and was built by using the shiny web application framework \citep{chang_shiny_2018} in R. Therein the user can choose to filter the data by several parameters such as chemical class, organism habitat, test duration and test endpoints amongst others. A detailed list with example entries of the possible filter parameters can be found in \ref{table:meta}. In order to derive single endpoints for an organism groups and a compound of the partly extensive test data, the \etoxbase{} aggregates the test results according to the chosen filters in two steps. Firstly the filtered test results are aggregated by the CAS number, the chosen taxon plus the duration and secondly only by the CAS number. The former can't be influenced by the user and calculates either the minimum or the median depending on the amount of results to aggregate (n <= 2: minimum, if n > 2: median). Thereof the second step calculates the minimum, the maximum, the median, the mean or the geometric mean as aggregate. Besides the aggregation some cleaning can also be performed. The \etoxbase{} allows to exclude test results exhibiting concentrations that are higher than the actual water solubility of the respective compound at \ang{20} C. In any case the user can choose to exclude outliers. The outliers are selected by excluding values that exceed lower (0.25) and the upper (0.75) quartile by 1.5 times the interquartile range. All the steps described in this section are executed at each click in the app which runs the function fun\_ec50filter\_aggregation.R (cf. table \ref{table:processing-scripts}). Fast, adequate and live data manipulation is possible thanks to the data.table package \citep{dowle_data.table_2018}.

@@

\begin{figure}
    %\includestandalone[]{tikz/test}
    \includestandalone[width=\textwidth]{tikz/organigram}
    %\includestandalone[width=\textwidth]{tikz/organigram}
    \caption{Organigram}
    \label{fig:organigram}
\end{figure}

@@

pH data is not really available for, temperature data either \ref{table:epa-variables}. therefore it is hard to evaluate single toxicities. We call for more concise reporting of ecotoxicological test data. It's all in articles but hardyl in a concise machine-readable format!

@@

\begin{sidewaystable}
    \csvautotabular[respect all]{data/scripts.csv}
    \caption{Processing scripts}
    \label{table:processing-scripts}
\end{sidewaystable}

@@

\bibliographystyle{apalike}
\bibliography{refs/references-etox-base}
